---
title: "BetaStability Quick Start Guide"
author: "Yu Gao(gaoyu19920914@gmail.com)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BetaStability Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# BetaStability Quick Start Guide

This vignette demonstrates the capabilities of the BetaStability package using the `varespec` and `varechem` datasets from the vegan package.

## Installation

First, install the package from GitHub:

```{r installation, eval = FALSE}
# Install from GitHub
# install.packages("devtools")
# devtools::install_github("gaoyu19920914/betaStability")

# OR install from BioConductor (in the future when it's available)
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("betaStability")
```

## Loading Required Packages

Load the BetaStability package and the vegan package for test data:

```{r load-packages}
library(betaStability)
library(vegan)
library(ggplot2)
```

## Loading Test Data

Load the `varespec` (community data) and `varechem` (environmental metadata) datasets from the vegan package:

```{r load-data}
data(varespec)
data(varechem)

# Inspect the data
head(varespec)
head(varechem)

# Dimensions of the datasets
cat("Dimensions of varespec:", dim(varespec), "\n")
cat("Dimensions of varechem:", dim(varechem), "\n")
```

## Calculating Stability with Single Method

Calculate stability using a single prediction method (`linearPred`):

```{r single-method}
# Calculate stability with linearPred
result_linear <- betaStability(comtable = varespec, 
                               envmeta = varechem, 
                               method = "linearPred")

# Inspect the result
head(result_linear)
length(result_linear)
```

## Calculating Stability with Multiple Methods

Calculate stability using multiple prediction methods:

```{r multiple-methods}
# Calculate stability with multiple methods
results_multi <- betaStability(comtable = varespec, 
                              envmeta = varechem, 
                              method = c("linearPred", "mlPred", "glmPred"))

# Inspect the result
head(results_multi)
dim(results_multi)
```

## Calculating Stability with All Methods

Calculate stability using all available prediction methods:

```{r all-methods}
# Calculate stability with all methods
results_all <- betaStability(comtable = varespec, 
                            envmeta = varechem, 
                            method = "all")

# Inspect the result
head(results_all)
dim(results_all)
colnames(results_all)
```

## Visualizing Stability Results

Visualize the stability results using the `plotStability` function:

```{r plot-single}
# Plot stability results for single method
p1 <- plotStability(result_linear)
p1
```

```{r plot-multi}
# Plot stability results for multiple methods
p2 <- plotStability(results_multi)
p2
```

```{r plot-all}
# Plot stability results for all methods
p3 <- plotStability(results_all)
p3
```

## Customizing Site Names

You can also customize the site names in the plot:

```{r custom-sitenames}
# Create custom site names
custom_sitenames <- paste("Site", 1:nrow(varespec))

# Plot with custom site names
p4 <- plotStability(results_multi, sitenames = custom_sitenames)
p4
```

## Conclusion

The BetaStability package provides a comprehensive framework for calculating and visualizing site stability using various prediction methods. Key features include:

1. **Multiple prediction methods**: Choose from linear, multiple linear model, generalized linear model, generalized additive model, generalized dissimilarity model, random forest, and xgboost models.
2. **Flexible input options**: Use precomputed distance matrices or let the package compute them automatically.
3. **Easy visualization**: The `plotStability` function creates informative plots of stability results.
4. **Convenient "all" method**: Run all prediction methods with a single parameter setting and receive a summary of results.

This package is designed to help researchers and ecologists assess site stability based on the beta stability of communities, providing valuable insights for ecosystem management.
